Data Preparation

### 2025-10-26 code/06_data_preparation.Rmd
Prisma checklist and diagram added, changes to figure code added more tables for supplementary material , updated reviewed code from MO and added proceedings csl file for bibliography.

---
title: "Data Preparation"
author: "Shreya D."
date: '`r Sys.Date()`'
output: html_document
license: CC-BY-4.0
email: shreya.dimri@uni-bielefeld.de
---

We have merged the data from different sources after data extraction, cleaned the typos and made it more usable. We have excluded the articles that did not pass full-text screening and the fitness proxies that we could not find a direction for the effect. These are steps done in the script **05_data_cleaning.Rmd** that generates the file **dataset_after_cleaning.csv**

In this script we will continue to prepare our dataset for use in meta-analysis, calculates variables for effect size estimation (SMDH, lnRR).

@@

data/03_data_cleaning/dataset_after_cleaning.csv

@@

```{r setup, warning=FALSE, message=FALSE, echo=F}

# Setting up workspace
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}
pacman::p_load(here, tidyverse, dplyr, stringr, metafor, orchaRd,ggpubr,ggrepel) #Packages needed

rm(list=ls()) ## cleaning up

set.seed(8341) #set seed for reproducibility (for random number generation)

# Loading cleaned dataset

dataset_after_cleaning <- read_csv(here::here("data/03_data_cleaning/dataset_after_cleaning.csv"))

## Now all the changes we make to the file.. we will store them in data_analysis

dataset_analysis<-dataset_after_cleaning
```

#### Loading functions

Some functions we will need in this manuscript.. sourced mainly from Nakagawa 2022 : Online tutorial <https://alistairmcnairsenior.github.io/Miss_SD_Sim/>

# Function to calculate Geary's "number"
  geary <- function(mean, sd, n){
    (1 / (sd / mean)) * ((4*n^(3/2)) / (1 + 4*n))
  }

# Shinichi Nakagawa, Daniel W. A. Noble, Malgorzata Lagisz, Rebecca Spake, Wolfgang Viechtbauer and Alistair M. Senior. 2022. A robust and readily implementable method for the meta-analysis of response ratios with and without missing standard deviations. Ecology Letters, DOI: 10.1111/ele.14144

  # Functions provided in the workalong
cv_avg <- function(x, sd, n, group, data, label = NULL, sub_b = TRUE, cv2=FALSE){

  # Check if the name is specified or not. If not, then assign it the name of the mean, x, variable input in the function. https://stackoverflow.com/questions/60644445/converting-tidyeval-arguments-to-string
  if(is.null(label)){
    label <- purrr::map_chr(enquos(x), rlang::as_label)
  }

  # Calculate between study CV. Take weighted mean CV within study, and then take a weighted mean across studies of the within study CV. Weighted based on sample size and pooled sample size.
  b_grp_cv_data <- data                                             %>%
    dplyr::group_by({{group}})                            %>%
    dplyr::mutate(   w_CV2 = weighted_CV({{sd}}, {{x}}, {{n}}, cv2=cv2),
                     n_mean = mean({{n}}, na.rm = TRUE))   %>%
    dplyr::ungroup(.)                                     %>%
    dplyr::mutate(b_CV2 = weighted.mean(w_CV2, n_mean, na.rm = TRUE), .keep = "used")

  # Make sure that label of the calculated columns is distinct from any other columns
  names(b_grp_cv_data) <- paste0(names(b_grp_cv_data), "_", label)

  # Append these calculated columns back to the original data and return the full dataset.
  if(sub_b){
    b_grp_cv_data <- b_grp_cv_data %>% dplyr::select(grep("b_", names(b_grp_cv_data)))
    dat_new <- cbind(data, b_grp_cv_data)
  } else {
    dat_new <- cbind(data, b_grp_cv_data)
  }

@@

# You also need the helper function

weighted_CV <- function(sd, x, n, cv2=FALSE){
  if(cv2){
    weighted.mean(na_if((sd / x)^2, Inf), n, na.rm = TRUE)
  }else{
    weighted.mean(na_if((sd / x), Inf), n, na.rm = TRUE)^2
  }
}

#' @title lnrr_laj
#' @description Calculates log response ratio based on Taylor expansion from Jajeunesse 2011
#' @param m1 Mean of treatment group 1
#' @param m2 Mean of treatment group 2
#' @param cv1_2 Coefficient of variation squared (CV^2) for treatment group 1
#' @param cv2_2 Coefficient of variation squared (CV^2) for treatment group 2
#' @param n1 Sample size for treatment group 1
#' @param n2 Sample size for treatment group 2
#' @param taylor A logical indicating whether to calculate point estimate with Taylor expansion.
#'
lnrr_laj <- function(m1, m2, cv1_2, cv2_2, n1, n2, taylor=TRUE){
  if(taylor){
    log(m1 / m2) + 0.5*((cv1_2 / n1) - (cv2_2 / n2))
  } else {
    log(m1 / m2)
  }
}

#' @title v_lnrr_laj
#' @description Calculates the sampling variance for log response ratio based on second order Taylor expansion proposed by Lajeunesse 2011
#' @param cv1_2 Coefficient of variation squared (CV^2) for treatment group 1
#' @param cv2_2 Coefficient of variation squared (CV^2) for treatment group 2
#' @param n1 Sample size for treatment group 1
#' @param n2 Sample size for treatment group 2
#' @param taylor A logical indicating whether to calculate point estimate with Taylor expansion.
v_lnrr_laj <- function(cv1_2, cv2_2, n1, n2,  taylor=TRUE){
  if(taylor){
  ((cv1_2) / n1) + ((cv2_2) / n2) +
    ((cv1_2)^2 / (2*n1^2)) + ((cv2_2)^2 / (2*n2^2))
  } else {
    ((cv1_2) / n1) + ((cv2_2) / n2)
  }
}

# Preparing Random Effects

@@

"(1) paper_ID, which encompasses estimates extracted from the same primary study,

(2) experiment_ID,

(3) group_ID

(4) repeated_traits_ID, and

(5) observation_ID which corresponds to a unit-level observation identity and models within-study variance

@@

-   We also need to add `observation ID` which is the unit-level identity (unique for each row) for within-study variance.

-   In the way we extracted this information, each paper_ID needs to be added to the random effect level of `experiment_ID`, `group_ID` and `repeated_trait_ID` to make sure they are unique values.

```{r Prepare Random Effects as Unique ID}

dataset_analysis<-dataset_analysis%>%

  mutate(Observation_ID = as.factor(row_number()))%>%

  mutate(experiment_ID_coded = as.factor(paste(paper_ID, experiment_ID, sep="_")),.after=experiment_ID)%>%
  mutate(group_ID_coded = as.factor(paste(paper_ID, group_ID, sep="_")),.after=group_ID)%>%
  mutate(repeated_trait_ID_coded = as.factor(paste(paper_ID,repeated_trait_ID, sep="_")),.after=repeated_trait_ID)

# random_effects_check<-dataset_analysis%>%select(paper_ID,experiment_ID,group_ID,repeated_trait_ID, experiment_ID_coded,group_ID_coded, repeated_trait_ID_coded)

-   We also decided to add `population ID` since we noticed several studies were performed in the same geographical location.

```{r}
dataset_analysis<-dataset_analysis%>%
  mutate(population_ID = as.factor(population_location))

# Let's look at the nammes of the locations
# unique(dataset_analysis$population_ID)

# There are clearly many repeated names, let's clean this up a bit

dataset_analysis <- dataset_analysis %>%
  mutate(
    population_ID = str_squish(as.character(population_location)),  # Clean whitespace
    population_ID = case_when(
      # Lodz
      str_detect(population_ID, regex("lodz|łódź|central poland", ignore_case = TRUE)) ~ "Lodz, Poland",

@@

      # for clean but unmatched names
      TRUE ~ str_to_title(population_ID)
    )
  )

# Effective Sample Size Calculation

@@

```{r effective sample size}

dataset_analysis<-dataset_analysis%>%
  mutate(effective_n_experiment=(n_experiment/shared_experiment),.after=n_experiment)%>%
  mutate(effective_n_control=(n_control/shared_control),.after=n_control)
```

@@

```{r}
dataset_analysis%>%
  summarise(control_min=min(effective_n_control,na.rm=T),
            control_max=max(effective_n_control,na.rm=T),
            na_control = sum(is.na(effective_n_control)),
            experiment_min=min(effective_n_experiment,na.rm=T),
            experiment_max=max(effective_n_experiment,na.rm=T),
             na_experiment = sum(is.na(effective_n_experiment)))
```

We have some NA and some values of effective sample size 1 and below.. lets see where this is

```{r}
low_sample<-dataset_analysis%>%
  filter(effective_n_control<2 | effective_n_experiment<2)
```

@@

# Converting variance to SD

@@

# For experimental group
dataset_analysis <- dataset_analysis %>%
  mutate(sd_experiment = if_else(type_measure_dispersion_experiment %in% c('se', 'SE'), #condition to check
      measure_dispersion_experiment * sqrt(as.numeric(n_experiment)), # do this if true
      measure_dispersion_experiment), #else do this
      .after=measure_dispersion_experiment)
# For control group
dataset_analysis <- dataset_analysis %>%
  mutate(sd_control = if_else(type_measure_dispersion_control %in% c('se', 'SE'), #condition to check
      measure_dispersion_control* sqrt(as.numeric(n_control)), # do this if true
      measure_dispersion_control), #else do this
      .after=measure_dispersion_control)
```

@@

# No effect size provided

@@

```{r}
# To calculate means by groups
trait_mean<- dataset_analysis%>%
  group_by(trait_type)%>%
  summarise(
    cumulative_mean = mean (
      c(measure_central_tendency_experiment,measure_central_tendency_control),
      na.rm=T
    ),
    count = n()
  )%>%
  arrange(desc(count))

# Now I need to assign the mean values from this to the missing ES proxies
dataset_analysis<-dataset_analysis%>%
  left_join(trait_mean, by= "trait_type")%>%
  mutate(measure_central_tendency_experiment = if_else((
    proxy_comment == "use 0 as ES" & !is.na(proxy_comment)),
    cumulative_mean,
    measure_central_tendency_experiment),
    measure_central_tendency_control = if_else((
    proxy_comment == "use 0 as ES" & !is.na(proxy_comment)),
    cumulative_mean,
    measure_central_tendency_control),)%>%
  select(-cumulative_mean)

# Effect Size Estimation

## Calculate lnRR

### Geary's Test

```{r}
# Assumption of normality assumed to be approximately correct when values are >= 3.
dataset_analysis<- dataset_analysis %>% 
         mutate(geary_control = geary(measure_central_tendency_control,
                                      sd_control, 
                                      n_control),
                    geary_trt = geary(measure_central_tendency_experiment,
                                      sd_experiment, 
                                      n_experiment),
                   geary_test = ifelse(geary_control >= 3 & geary_trt >= 3, "pass", "fail"))

# How many fail?
    geary_res <-dataset_analysis %>% group_by(geary_test) %>% summarise(n = n()) %>%  data.frame()

@@

```{r for lnRR}

## ─────────────────────────────────────────────────────────────
## 1.  Keeping rows where lnRR can be calculated. 
## Inferential stats and contingency table can only be used for SMDH.
##    — store the dropped rows to inspect and join back later
## ─────────────────────────────────────────────────────────────

only_SMDH <-dataset_analysis%>%
  filter(proxy_decision == "contingency table"
         |proxy_decision == "SMDH only")

@@

## -------------------------------------------------------------------------
## 2. Rows whose means are > 0 (lnRR needs ratio scale data)
##  - store the dropped rows to inspect and join back later to the dataset
## -------------------------------------------------------------------------

neg_means<- dataset_analysis%>%
 filter(measure_central_tendency_experiment <= 0 |
         measure_central_tendency_control <= 0)

@@

## -------------------------------------------------------------------------
## 3. Estimating lnRR using metafor::escalc for cases where SD is available
## -------------------------------------------------------------------------

# Using function 'escalc()' (with measure=ROM and 'vtype=LS')

dataset_analysis<- escalc(measure = "ROM", 
              vtype= "LS",
                        n1i = effective_n_experiment, 
                        n2i = effective_n_control,
                        m1i = measure_central_tendency_experiment, 
                        m2i = measure_central_tendency_control,
                        sd1i = sd_experiment, 
                        sd2i = sd_control,
                        data = dataset_analysis, 
                        var.names = c('lnRR',
                                      'lnRR_variance'),
                        add.measure = FALSE,
                        append = TRUE)

@@

not_calculated<-dataset_analysis%>%filter(is.na(lnRR)|is.na(lnRR_variance))

@@

## -------------------------------------------------------------------------
## 4. Estimating lnRR using Missing Case approach from Nakagawa 2022
##    see https://alistairmcnairsenior.github.io/Miss_SD_Sim/
## -------------------------------------------------------------------------

@@

dataset_analysis<-dataset_analysis%>%
  mutate(lnRR = if_else((
    proxy_comment == "use 0 as ES" & !is.na(proxy_comment)),
    lnrr_laj,
    lnRR),
   lnRR_variance = if_else((
    proxy_comment == "use 0 as ES" & !is.na(proxy_comment)),
    v_lnrr_laj,
    lnRR_variance),)%>%
select(-b_CV2_exp,-b_CV2_con,-cv_Control,-cv_Experimental,-cv2_con_new,-cv2_exp_new,-lnrr_laj,-v_lnrr_laj)

## How many lnRR were not calculated now?

not_calculated<-dataset_analysis%>%filter(is.na(lnRR)|is.na(lnRR_variance))

@@

dataset_analysis<-bind_rows(dataset_analysis,only_SMDH,neg_means)
```

@@

```{r correct sign for ES}

dataset_analysis<-dataset_analysis%>%
  mutate(lnRR_sign=lnRR*proxies_sign)
```

### Outliers Check

@@

```{r Funnel Plot}

funnel_lnRR <- dataset_analysis %>%
  ggplot(aes(x = lnRR_sign, y = 1 / lnRR_variance)) +
  geom_point(color= "#7DCE82", alpha = 0.4, size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#8B4513", linewidth=0.65) +  # Add vertical dashed line at x = 0
  ggtitle("Funnel Plot for lnRR") +
  xlab("lnRR") +
  ylab("Precision (1/Variance)")

@@

rows_with_missing <- dataset_analysis %>%
  filter(is.na(lnRR_sign) | is.na(lnRR_variance))

@@

## Calculate SMDH

-   For results only reported as inferential statistics (e.g., t-test, chi-square), we will use the equations provided by (Lajeunesse 2013; Nakagawa and Cuthill 2007; and other sources if needed) to calculate their corresponding SMD estimates for the analyses.

-   Note that lnRR cannot be calculated from inferential statistics and that only SMD without heteroskedasticity correction can be calculated from inferential statistics.

```{r}
#-------------------------------------------------------------------------
# 1. SMDH for most cases where mean, variance and sample size was reported
#-------------------------------------------------------------------------

@@

dataset_analysis<- escalc(measure = "SMD", 
              vtype= "LS",
                        n1i = effective_n_experiment, 
                        n2i = effective_n_control,
                        m1i = measure_central_tendency_experiment, 
                        m2i = measure_central_tendency_control,
                        sd1i = sd_experiment, 
                        sd2i = sd_control,
                        data = dataset_analysis, 
                        var.names = c('SMDH',
                                      'SMDH_variance'),
                        add.measure = FALSE,
                        append = TRUE)

# Cases where it was not calculated using mean, variance and sample size
noSMDH<-dataset_analysis%>%filter(is.na(SMDH))

#-------------------------------------------------------------------------
# 2. No effect of the treatment is present but 
# estimates (means, test statistics or effect size) were not reported
#-------------------------------------------------------------------------

@@

dataset_analysis <- dataset_analysis %>%
  mutate(SMDH = if_else(proxy_comment == "use 0 as ES" & !is.na(proxy_comment), (measure_central_tendency_experiment-measure_central_tendency_control), SMDH))%>%
  mutate(SMDH_variance = if_else(proxy_comment == "use 0 as ES" & !is.na(proxy_comment), ((effective_n_experiment+effective_n_control)/(effective_n_experiment*effective_n_control)), SMDH_variance))

# Cases where it was not calculated using mean, variance and sample size
noSMDH<-dataset_analysis%>%filter(is.na(SMDH))

@@

```{r}
contingency_table<-dataset_analysis%>%filter(proxy_decision=="contingency table")

@@

contingency_table <- contingency_table %>%
  mutate(
    # Calculate the number of successes and failures for experiment and control groups
    success_experiment = round((measure_central_tendency_experiment / 100) * n_experiment),
    failure_experiment = n_experiment - success_experiment,
    success_control = round((measure_central_tendency_control / 100) * n_control),
    failure_control = n_control - success_control)
    # Calculate Log Odds Ratio (lnOR)
contingency_table<- escalc(measure="OR", ai=success_experiment, bi=failure_experiment, ci=success_control, di=failure_control, data=contingency_table, var.names = c('lnOR',
                                      'lnOR_variance'))
# Calculate SMDH
contingency_table <- contingency_table %>%
  mutate(SMDH=(lnOR*sqrt(3))/pi,
         SMDH_variance=(3*lnOR_variance)/pi^2)

# Update SMDH and SMDH_variance based on Observation_ID
dataset_analysis <- rows_update(
  x = dataset_analysis,           # Target dataset to update
  y = contingency_table %>% 
        select(Observation_ID, SMDH, SMDH_variance), # Updates from contingency_table
  by = "Observation_ID"       # Match on Observation_ID
)
```

@@

![From Cooper2019 Table 11.3](/images/ancova-d.png)

@@

![From Cooper2019 Table 11.3](/images/anova-d.png)

@@

ancova_SMDH<-dataset_analysis%>%filter(statistics_type=="ANCOVA")
# 3 effect size where SMDH needs to be estimated from ANCOVA

@@

ancova_SMDH<-ancova_SMDH%>%
  mutate(SMDH = sign_relationship*(d_anova(statistics_value,n_experiment,n_control)))%>%
    mutate(SMDH_variance = var_d_anova(SMDH, n_experiment,n_control))

>>>>

---
title: "Data Preparation"
author: "Shreya D."
date: '`r Sys.Date()`'
output: html_document
license: CC-BY-4.0
email: shreya.dimri@uni-bielefeld.de
---
## Preface

In this script we will continue to prepare our dataset for use in meta-analysis and calculate variables for effect size estimation (SMDH, lnRR). 
In the previous script [05_data_cleaning.Rmd](https://github.com/shreyadimri/Green_Nest_Material/blob/main/code/05_data_cleaning.Rmd) we merged the data from different sources after data extraction, and performed several data cleaning procedures, including the standardization of terms. We have excluded the articles that did not pass full-text screening and the fitness proxies that we could not find a direction for the effect. The resulting data were saved as [dataset_after_cleaning.csv](https://github.com/shreyadimri/Green_Nest_Material/blob/main/data/03_data_cleaning/dataset_after_cleaning.csv)

@@

[data/03_data_cleaning/dataset_after_cleaning.csv](https://github.com/shreyadimri/Green_Nest_Material/blob/main/data/03_data_cleaning/dataset_after_cleaning.csv)

@@

* Install and load required packages

```{r setup, warning=FALSE, message=FALSE}
## Setting up workspace using pacman
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

## installing orchaRd from GitHub
if (!requireNamespace("orchaRd", quietly = TRUE)) {
## installing orchaRd from GitHub
pacman::p_install_gh("daniel1noble/orchaRd", ref = 'main')
}

## Install/load required packages
pacman::p_load(
  here,
  tidyverse,
  metafor,
  orchaRd,
  ggpubr)
```

* Load cleaned dataset [dataset_after_cleaning.csv](https://github.com/shreyadimri/Green_Nest_Material/blob/main/data/03_data_cleaning/dataset_after_cleaning.csv)

```{r load cleaned data}
# set seed for reproducibility (for random number generation)
set.seed(8341)

# Loading cleaned dataset
dataset_after_cleaning <- read_csv(here::here("data/03_data_cleaning/dataset_after_cleaning.csv"))

## create working copy to store changes
dataset_analysis <- dataset_after_cleaning
```

#### Custom functions

For the presented analyses, several functions are defined, most of them defined in [Nakagawa, Noble, Lagisz, Spake, Viechtbauer, & Senior (2023)](https://doi.org/10.1111/ele.14144) and the accompanying online supplement <https://alistairmcnairsenior.github.io/Miss_SD_Sim/>

```{r define custom functions}
## Function to calculate Geary's "number"
geary <- function(mean, sd, n){
  (1 / (sd / mean)) * ((4*n^(3/2)) / (1 + 4*n))
}

## Functions provided in the workalong
cv_avg <- function(x, sd, n, group, data, label = NULL, sub_b = TRUE, cv2 = FALSE) {

## Check if the name is specified or not. If not, then assign it the name of the mean, x,
## variable input in the function.
##  https://stackoverflow.com/questions/60644445/converting-tidyeval-arguments-to-string
  if (is.null(label)) {
    label <- purrr::map_chr(enquos(x), rlang::as_label)
  }

  ## Calculate between study CV. 
  ## Take weighted mean CV within study, and then take a weighted mean across studies 
  ## of the within study CV. Weighted based on sample size and pooled sample size.
  b_grp_cv_data <- data %>%
    dplyr::group_by({{group}}) %>%
    dplyr::mutate(   w_CV2 = weighted_CV({{sd}}, {{x}}, {{n}}, cv2 = cv2),
                     n_mean = mean({{n}}, na.rm = TRUE)) %>%
    dplyr::ungroup(.) %>%
    dplyr::mutate(b_CV2 = weighted.mean(w_CV2, n_mean, na.rm = TRUE), .keep = "used")

  ## Make sure that label of the calculated columns is distinct from any other columns
  names(b_grp_cv_data) <- paste0(names(b_grp_cv_data), "_", label)

  ## Append these calculated columns back to the original data and return the full dataset.
  if(sub_b){
    b_grp_cv_data <- b_grp_cv_data %>%
      dplyr::select(grep("b_", names(b_grp_cv_data)))
    dat_new <- cbind(data, b_grp_cv_data)
  } else {
    dat_new <- cbind(data, b_grp_cv_data)
  }

@@

## You also need the helper function
weighted_CV <- function(sd, x, n, cv2 = FALSE) {
  if(cv2) {
    weighted.mean(na_if((sd / x)^2, Inf), n, na.rm = TRUE)
  } else {
    weighted.mean(na_if((sd / x), Inf), n, na.rm = TRUE)^2
  }
}

#' @title lnrr_laj
#' @description Calculates log response ratio based on Taylor expansion from Jajeunesse 2011
#' @param m1 Mean of treatment group 1
#' @param m2 Mean of treatment group 2
#' @param cv1_2 Coefficient of variation squared (CV^2) for treatment group 1
#' @param cv2_2 Coefficient of variation squared (CV^2) for treatment group 2
#' @param n1 Sample size for treatment group 1
#' @param n2 Sample size for treatment group 2
#' @param taylor A logical indicating whether to calculate point estimate with Taylor expansion.
#'
lnrr_laj <- function(m1, m2, cv1_2, cv2_2, n1, n2, taylor = TRUE) {
  if(taylor) {
    log(m1 / m2) + 0.5*((cv1_2 / n1) - (cv2_2 / n2))
  } else {
    log(m1 / m2)
  }
}

#' @title v_lnrr_laj
#' @description Calculates the sampling variance for log response ratio based on second order Taylor expansion proposed by Lajeunesse 2011
#' @param cv1_2 Coefficient of variation squared (CV^2) for treatment group 1
#' @param cv2_2 Coefficient of variation squared (CV^2) for treatment group 2
#' @param n1 Sample size for treatment group 1
#' @param n2 Sample size for treatment group 2
#' @param taylor A logical indicating whether to calculate point estimate with Taylor expansion.
v_lnrr_laj <- function(cv1_2, cv2_2, n1, n2,  taylor = TRUE) {
  if(taylor){
    ((cv1_2) / n1) + ((cv2_2) / n2) +
      ((cv1_2)^2 / (2*n1^2)) + ((cv2_2)^2 / (2*n2^2))
  } else {
    ((cv1_2) / n1) + ((cv2_2) / n2)
  }
}
```

## Preparing Random Effects

@@

* (1) paper_ID, which encompasses estimates extracted from the same primary study,

* (2) experiment_ID,

* (3) group_ID

* (4) repeated_traits_ID, and

* (5) observation_ID which corresponds to a unit-level observation identity and models within-study variance

@@

* We also need to add `observation ID` which is the unit-level identity (unique for each row) for within-study variance.

* In the way we extracted this information, each paper_ID needs to be added to the random effect level of `experiment_ID`, `group_ID` and `repeated_trait_ID` to make sure they are unique values.

```{r Prepare Random Effects as Unique ID}
dataset_analysis <- dataset_analysis %>%
  mutate(Observation_ID = as.factor(row_number())) %>%
  mutate(experiment_ID_coded = as.factor(paste(paper_ID, experiment_ID, sep="_")),
         .after = experiment_ID) %>%
  mutate(group_ID_coded = as.factor(paste(paper_ID, group_ID, sep="_")),
         .after = group_ID) %>%
  mutate(repeated_trait_ID_coded = as.factor(paste(paper_ID,repeated_trait_ID, sep="_")),
         .after = repeated_trait_ID)

# random_effects_check<-dataset_analysis%>%select(paper_ID,experiment_ID,group_ID,repeated_trait_ID, experiment_ID_coded,group_ID_coded, repeated_trait_ID_coded)
```

* We also decided to add `population ID` since we noticed several studies were performed in the same geographical location.

```{r cleaning location names}
dataset_analysis <- dataset_analysis %>%
  mutate(population_ID = as.factor(population_location))

# Let's look at the names of the locations
# unique(dataset_analysis$population_ID)
# There are clearly many repeated names, let's clean this up a bit
dataset_analysis <- dataset_analysis %>%
  mutate(
    population_ID = str_squish(as.character(population_location)),  # Clean whitespace
    population_ID = case_when(
      # Lodz
      str_detect(population_ID, regex("lodz|łódź|central poland", ignore_case = TRUE)) ~ "Lodz, Poland",

@@

      # for clean but unmatched names
      TRUE ~ str_to_title(population_ID)
    )
  )
```

## Effective Sample Size Calculation

@@

```{r effective sample size}
dataset_analysis <- dataset_analysis %>%
  mutate(effective_n_experiment = (n_experiment/shared_experiment), .after = n_experiment) %>%
  mutate(effective_n_control = (n_control/shared_control),.after = n_control)
```

@@

```{r}
dataset_analysis %>%
  summarise(control_min = min(effective_n_control, na.rm = T),
            control_max = max(effective_n_control, na.rm = T),
            na_control = sum(is.na(effective_n_control)),
            experiment_min = min(effective_n_experiment, na.rm = T),
            experiment_max = max(effective_n_experiment, na.rm = T),
            na_experiment = sum(is.na(effective_n_experiment)))
```

We have some NA and some values of effective sample size 1 and below lets see where this is

```{r filter out low sample size}
low_sample <- dataset_analysis %>%
  filter(effective_n_control < 2 | effective_n_experiment < 2)
```

@@

## Converting variance to SD

@@

# For experimental group
dataset_analysis <- dataset_analysis %>%
  mutate(sd_experiment = if_else(
    condition = type_measure_dispersion_experiment %in% c('se', 'SE'), 
    true = measure_dispersion_experiment * sqrt(as.numeric(n_experiment)), 
    false = measure_dispersion_experiment), 
    .after = measure_dispersion_experiment)

# For control group
dataset_analysis <- dataset_analysis %>%
  mutate(sd_control = if_else(#
    condition = type_measure_dispersion_control %in% c('se', 'SE'), 
    true = measure_dispersion_control* sqrt(as.numeric(n_control)), 
    false = measure_dispersion_control),
    .after=measure_dispersion_control)
```

@@

## No effect size provided

@@

```{r}
# To calculate means by groups
trait_mean<- dataset_analysis %>%
  group_by(trait_type) %>%
  summarise(cumulative_mean = mean (
    c(measure_central_tendency_experiment,measure_central_tendency_control),
    na.rm = T
    ),
    count = n()
  ) %>%
  arrange(desc(count))

# Now I need to assign the mean values from this to the missing ES proxies
dataset_analysis <- dataset_analysis %>%
  left_join(trait_mean, by = "trait_type") %>%
  mutate(measure_central_tendency_experiment = if_else((
    proxy_comment == "use 0 as ES" & !is.na(proxy_comment)),
    cumulative_mean,
    measure_central_tendency_experiment),
    measure_central_tendency_control = if_else((
      proxy_comment == "use 0 as ES" & !is.na(proxy_comment)),
      cumulative_mean,
      measure_central_tendency_control),) %>%
  select(-cumulative_mean)
```

## Effect Size Estimation

### Calculate lnRR

#### Geary's Test

```{r}
# Assumption of normality assumed to be approximately correct when values are >= 3.
dataset_analysis <- dataset_analysis %>% 
  mutate(geary_control = geary(measure_central_tendency_control,
                               sd_control, 
                               n_control),
         geary_trt = geary(measure_central_tendency_experiment,
                           sd_experiment, 
                           n_experiment),
         geary_test = ifelse(geary_control >= 3 & geary_trt >= 3, "pass", "fail"))

# How many fail?
geary_res <- dataset_analysis %>%
  group_by(geary_test) %>% 
  summarise(n = n()) %>%  
  data.frame()

@@

```{r for lnRR}
## 1.  Keeping rows where lnRR can be calculated. 
## Inferential stats and contingency table can only be used for SMDH.
##    — store the dropped rows to inspect and join back later

only_SMDH <-dataset_analysis %>%
  filter(proxy_decision == "contingency table"
         |proxy_decision == "SMDH only")

@@

## 2. Rows whose means are > 0 (lnRR needs ratio scale data)
##  - store the dropped rows to inspect and join back later to the dataset

neg_means <- dataset_analysis %>%
  filter(measure_central_tendency_experiment <= 0 |
           measure_central_tendency_control <= 0)

@@

## 3. Estimating lnRR using metafor::escalc for cases where SD is available
# Using function 'escalc()' (with measure=ROM and 'vtype=LS')

dataset_analysis <- escalc(
  measure = "ROM", 
  vtype = "LS",
  n1i = effective_n_experiment, 
  n2i = effective_n_control,
  m1i = measure_central_tendency_experiment, 
  m2i = measure_central_tendency_control,
  sd1i = sd_experiment, 
  sd2i = sd_control,
  data = dataset_analysis, 
  var.names = c('lnRR',
                'lnRR_variance'),
  add.measure = FALSE,
  append = TRUE)

@@

not_calculated <- dataset_analysis %>% 
  filter(is.na(lnRR)|is.na(lnRR_variance))

@@

## 4. Estimating lnRR using Missing Case approach from Nakagawa 2022
##    see https://alistairmcnairsenior.github.io/Miss_SD_Sim/

@@

dataset_analysis <- dataset_analysis %>%
  mutate(lnRR = if_else((
    proxy_comment == "use 0 as ES" & !is.na(proxy_comment)),
    lnrr_laj,
    lnRR),
    lnRR_variance = if_else((
      proxy_comment == "use 0 as ES" & !is.na(proxy_comment)),
      v_lnrr_laj,
      lnRR_variance),)%>%
  select(-b_CV2_exp, -b_CV2_con, -cv_Control, -cv_Experimental,
         -cv2_con_new, -cv2_exp_new, -lnrr_laj, -v_lnrr_laj)

## How many lnRR were not calculated now?
not_calculated <- dataset_analysis %>% 
  filter(is.na(lnRR)|is.na(lnRR_variance))

@@

dataset_analysis <- bind_rows(dataset_analysis, only_SMDH, neg_means)
```

#### Sign for Effect Size

@@

```{r correct sign for ES}
dataset_analysis <- dataset_analysis %>%
  mutate(lnRR_sign = lnRR*proxies_sign)
```

#### Outliers Check

@@

```{r Funnel Plot}
funnel_lnRR <- dataset_analysis %>%
  ggplot(aes(x = lnRR_sign, y = 1 / lnRR_variance)) +
  geom_point(color= "#7DCE82", alpha = 0.4, size = 3) +
  # Add vertical dashed line at x = 0
  geom_vline(xintercept = 0, linetype = "dashed", color = "#8B4513", linewidth=0.65) +  
  ggtitle("Funnel Plot for lnRR") +
  xlab("lnRR") +
  ylab("Precision (1/Variance)")

@@

rows_with_missing <- dataset_analysis %>%
  filter(is.na(lnRR_sign) | is.na(lnRR_variance))
```

@@

### Calculate SMDH

* For results only reported as inferential statistics (e.g., t-test, chi-square), we will use the equations provided by (Lajeunesse 2013; Nakagawa and Cuthill 2007; and other sources if needed) to calculate their corresponding SMD estimates for the analyses.

* Note that lnRR cannot be calculated from inferential statistics and that only SMD without heteroskedasticity correction can be calculated from inferential statistics.

```{r}
# 1. SMDH for most cases where mean, variance and sample size was reported
#-------------------------------------------------------------------------

@@

dataset_analysis <- escalc(
  measure = "SMD", 
  vtype= "LS",
  n1i = effective_n_experiment, 
  n2i = effective_n_control,
  m1i = measure_central_tendency_experiment, 
  m2i = measure_central_tendency_control,
  sd1i = sd_experiment, 
  sd2i = sd_control,
  data = dataset_analysis, 
  var.names = c('SMDH',
                'SMDH_variance'),
  add.measure = FALSE,
  append = TRUE)

# Cases where it was not calculated using mean, variance and sample size
noSMDH <- dataset_analysis %>% 
  filter(is.na(SMDH))

# 2. No effect of the treatment is present but 
# estimates (means, test statistics or effect size) were not reported
#-------------------------------------------------------------------------

@@

dataset_analysis <- dataset_analysis %>%
  mutate(SMDH = if_else(proxy_comment == "use 0 as ES" & !is.na(proxy_comment), (measure_central_tendency_experiment-measure_central_tendency_control), SMDH)) %>%
  mutate(SMDH_variance = if_else(proxy_comment == "use 0 as ES" & !is.na(proxy_comment), ((effective_n_experiment+effective_n_control)/(effective_n_experiment*effective_n_control)), SMDH_variance))

# Cases where it was not calculated using mean, variance and sample size
noSMDH <- dataset_analysis %>% filter(is.na(SMDH))
```

@@

```{r contingency_table}
contingency_table <- dataset_analysis %>%
  filter(proxy_decision == "contingency table")

@@

contingency_table <- contingency_table %>%
  mutate(
    # Calculate the number of successes and failures for experiment and control groups
    success_experiment = round((measure_central_tendency_experiment / 100) * n_experiment),
    failure_experiment = n_experiment - success_experiment,
    success_control = round((measure_central_tendency_control / 100) * n_control),
    failure_control = n_control - success_control)

# Calculate Log Odds Ratio (lnOR)
contingency_table <- escalc(
  measure = "OR",
  ai = success_experiment, 
  bi = failure_experiment, 
  ci = success_control, 
  di = failure_control, 
  data = contingency_table,
  var.names = c('lnOR', 'lnOR_variance'))
# Calculate SMDH
contingency_table <- contingency_table %>%
  mutate(SMDH = (lnOR*sqrt(3))/pi,
         SMDH_variance = (3*lnOR_variance)/pi^2)

# Update SMDH and SMDH_variance based on Observation_ID
dataset_analysis <- rows_update(
  # Target dataset to update
  x = dataset_analysis,           
  y = contingency_table %>% 
    select(Observation_ID, SMDH, SMDH_variance), # Updates from contingency_table
  by = "Observation_ID"       # Match on Observation_ID
)
```

@@

![From Cooper2019 Table 11.3](../images/ancova-d.png)

@@

![From Cooper2019 Table 11.3](../images/anova-d.png)

@@

```{r}
ancova_SMDH <- dataset_analysis %>% 
  filter(statistics_type=="ANCOVA")
# 3 effect size where SMDH needs to be estimated from ANCOVA

@@

ancova_SMDH <- ancova_SMDH %>%
  mutate(SMDH = sign_relationship*(d_anova(statistics_value, n_experiment, n_control))) %>%
  mutate(SMDH_variance = var_d_anova(SMDH, n_experiment,n_control))
